# install required packages/library
```{r}
install.packages("matlib")
install.packages("rsample")
```

# import required packages/library
```{r}
library(matlib)
library(ggplot2)
library(rsample)
library(corrplot)
```

# load features dataset
```{r}
features<-as.matrix(read.csv(file="data/features.csv",header=F))
colnames(features)<-c("x1", "x3", "x4", "x5")
head(features)
```

# load target dataset
```{r}
target<-as.matrix(read.csv(file="data/target.csv",header=F))
colnames(target)<-c("X2")
head(target)
```

# load time series dataset
```{r}
time<-as.matrix((read.csv(file="data/timeseries.csv",header=F)))
colnames(time)<-c("T1")
head(time)
```

# Task 1: Preliminary Data Analysis
# time series of input signals
```{r}
features.ts<-ts(features,start=c(min(time),max(time)),frequency=1)
plot(features.ts,main="Time Series - Input Signal",xlab="Time",ylab="Input Signal")
```

# time series of output signal
```{r}
target.ts<-ts(target,start=c(min(time),max(time)),frequency=1)
plot(target.ts,main="Time Series - Output Signal",xlab="Time",ylab="Output Signal")
```

# Task 1.2: Distribution of Each Signal
```{r}
# histogram for temperature (x1)
density_x1<-density(features[,"x1"])
hist(features[,"x1"],freq=FALSE, main="Distribution of Temperature (x1)", xlab="Ambient temperature (°C)")
lines(density_x1, lwd=2)
```

```{r}
# histogram for ambient pressure (x3)
density_x3<-density(features[,"x3"])
hist(features[,"x3"],freq=FALSE, main="Distribution of Ambient Pressure (x3)", xlab="Atmospheric pressure (millibar)")
lines(density_x3, lwd=2)
```

```{r}
# histogram for relative humidity (x4)
density_x4<-density(features[,"x4"])
hist(features[,"x4"],freq=FALSE, main="Distribution of Relative Humidity (x4)", xlab="Humidity level (%)")
lines(density_x4, lwd=2)
```

```{r}
# histogram for exhaust vacuum (x5)
density_x5<-density(features[,"x5"])
hist(features[,"x5"],freq=FALSE, main="Distribution of Exhaust Vacuum (x5)", xlab="Vacuum collected from the steam turbine (cm Hg)")
lines(density_x5, lwd=2)
```

```{r}
# histogram for net hourly electrical energy output (X2)
density_x2<-density(target[,"X2"])
hist(target[,"X2"],freq=FALSE, main="Distribution of Net hourly electrical energy output (X2)", xlab="Net hourly electrical energy output in MW")
lines(density_x2, lwd=2)
```

# Task 1.3: Correlation and Scatter Plots
```{r}
# correlation
combined <- cbind(features, target)
correlation<-cor(combined)
print(correlation)
corrplot(correlation)
```

# scatter plot of x1 and X2
```{r}
plot(features[,"x1"], 
     target[,"X2"], 
     main="Temperature vs Net Hourly Electrical Energy Output", 
     xlab="Temperature (°C)", 
     ylab="Net Hourly Electrical Energy Output (MW)")
```

# scatter plot of x3 and X2
```{r}
plot(features[,"x3"], 
     target[,"X2"], 
     main="Ambient Pressure vs Net Hourly Electrical Energy Output", 
     xlab="Ambient Pressure (millibar)", 
     ylab="Net Hourly Electrical Energy Output (MW)")
```

# scatter plot of x4 and X2
```{r}
plot(features[,"x4"], 
     target[,"X2"], 
     main="Relative Humidity vs Net Hourly Electrical Energy Output", 
     xlab="Relative Humidity (%)", 
     ylab="Net Hourly Electrical Energy Output (MW)")
```

# scatter plot of x5 and X2
```{r}
plot(features[,"x5"], 
     target[,"X2"], 
     main="Exhaust Vacuum vs Net Hourly Electrical Energy Output", 
     xlab="Exhaust Vacuum (cm Hg)", 
     ylab="Net Hourly Electrical Energy Output (MW)")
```

# Task 2: Regression - Modelling the Between the Gene Expression
```{r}
# candidate nonlinear polynomial regression models
model1<-function(X, theta){
  theta[1]*X$x4 + theta[2]*X$x3^2 + theta[3]
}

model2<-function(X, theta){
  theta[1]*X$x4 + theta[2]*X$x3^2 + theta[3]*X$x5 + theta[4]
}

model3<-function(X, theta){
  theta[1]*X$x3 + theta[2]*X$x4 + theta[3]*X$x5^3
}

model4<-function(X, theta){
  theta[1]*X$x4 + theta[2]*X$x3^2 + theta[3]*X$x5^3 + theta[4]
}

model5<-function(X, theta){
  theta[1]*X$x4 + theta[2]*X$x1^2 + theta[3]*X$x3^2 + theta[4]
}
```

